---
title: "CollapseWindows"
output: html_document
date: "2025-04-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

This chunk generates data like we collected
```{r generate breakpoints}
n=5 #number of breakpoints simulated for this inversion
# assume each breakpoint start point could be within 500 bp of each other
# assume each CI is between 50 and 500 bases long

# inversion 1
set.seed(20934580)
inv_startBreakPt_lowerCI <- round(runif(n, 50, 500),0)
set.seed(94389523)
inv_startBreakPt_upperCI <- round(inv_startBreakPt_lowerCI + runif(n,100, 500),0)
set.seed(34853453)
inv_endBreakPt_lowerCI <- round(runif(n, 1000, 1500),0)
set.seed(54692349)
inv_endBreakPt_upperCI <- round(inv_endBreakPt_lowerCI + runif(n,100, 500),0)

# inversion 2
set.seed(5767871)
inv_startBreakPt_lowerCI2 <- round(runif(n, 3000, 3500),0)
set.seed(7108100)
inv_startBreakPt_upperCI2 <- round(inv_startBreakPt_lowerCI2 + runif(n,100, 500),0)
set.seed(9054960)
inv_endBreakPt_lowerCI2 <- round(runif(n, 4500, 5000),0)
set.seed(8137034)
inv_endBreakPt_upperCI2 <- round(inv_endBreakPt_lowerCI2 + runif(n,100, 500),0)

invdat1 <- data.frame(inv_startBreakPt_lowerCI,
                    inv_startBreakPt_upperCI,
                    inv_endBreakPt_lowerCI,
                    inv_endBreakPt_upperCI,
                    chrom = 1)
invdat2 <- data.frame(inv_startBreakPt_lowerCI=inv_startBreakPt_lowerCI2,
                    inv_startBreakPt_upperCI=inv_startBreakPt_upperCI2,
                    inv_endBreakPt_lowerCI=inv_endBreakPt_lowerCI2,
                    inv_endBreakPt_upperCI=inv_endBreakPt_upperCI2,
                    chrom = 1)

invdat3 <- data.frame(inv_startBreakPt_lowerCI=inv_startBreakPt_lowerCI+10,
                    inv_startBreakPt_upperCI=inv_startBreakPt_upperCI+10,
                    inv_endBreakPt_lowerCI=inv_endBreakPt_lowerCI+10,
                    inv_endBreakPt_upperCI=inv_endBreakPt_upperCI+10,
                    chrom = 2)
invdat <- rbind(invdat1, invdat2, invdat3)
rm(invdat1, invdat2, invdat3)

# make up some more data
invdat$mergeKey <- replicate(nrow(invdat),paste(sample(letters, 8), collapse=""))
invdat$method <- sample(c("methodA", "methodB"),nrow(invdat), replace=TRUE)
invdat$indID <- sample(c("ind1", "ind2","ind3","ind4"),nrow(invdat), replace=TRUE)
```


```{r plotCI}
plotCI <- function(lowerCI,upperCI,chrom, color="red"){
  if(sum(!(upperCI > lowerCI))==0){ # check the upper is greater than the lower
   dat <- data.frame(lowerCI, upperCI)
   dat2 <- dat[order(chrom,lowerCI),]
   plot(x=c(dat2$lowerCI, dat2$upperCI),
      y=rep(1:nrow(dat),2), col=color, ylab="breakpoint call", xlab="bp")
   arrows(dat2$lowerCI, 1:nrow(dat),
       dat2$upperCI,1:nrow(dat), col=color)
  }else{
    print("plotCI Error: At least one Upper CI is not larger than the corresponding Lower CI")
  } #end if
  return(dat2)
} # end function
```

```{r apply plotCI}
par(mfrow=c(2,1), mar=c(4,4,1,1))
plotCI(invdat$inv_startBreakPt_lowerCI, 
       invdat$inv_startBreakPt_upperCI, invdat$chrom)
plotCI(invdat$inv_endBreakPt_lowerCI, 
       invdat$inv_endBreakPt_upperCI, col="blue",
       invdat$chrom)
```



chrom <- invdat$chrom
lowerCI <- invdat$inv_startBreakPt_lowerCI
upperCI <- invdat$inv_startBreakPt_upperCI

```{r function collapseWindow}

collapseWindow <- function(lowerCI,upperCI, 
                           chrom,
                           mergeKey=NA, windowIDstartNum = 1){
  
  # check the upper is greater than the lower
  if(sum(!(upperCI > lowerCI))!=0){ 
    print("Error in collapseWindow(): At least one Upper CI is not larger than the corresponding Lower CI")
    break()
  }# end if
  
  # check that chromosome is supplied
  if (length(chrom)==0){
        print("Error in collapseWindow(): You must supply a chromosome or scaffold associated with each window")
    break()
  } #end if
  
  dat <- data.frame(lowerCI, upperCI, chrom, mergeKey)
   
  # First we sort the windows by their starting points
  dat2 <- dat[order(chrom,lowerCI),]
  dat2$windowID <- NA # add a column to store the windowID
  
   ########
  # Set the current extent of the CI
  # Assign the first windowID to the first window
  currentWindowID <- 1 #ID that you'd like to start with
  dat2$windowID[1] <- 1
  currentLowerCI <- dat2$lowerCI[1]
  currentUpperCI <- dat2$upperCI[1]
  
  ########
  # Create a window DF that will store the window info
  windowDF <- data.frame(windowID=NA, windowLowerCI=NA,
                         windowUpperCI = NA)
  windowDF[1,] <- c(currentWindowID, currentLowerCI,
                    currentUpperCI)
  ########
  ########
  # Loop through windows
  for (j in 2:nrow(dat2)){

     # make sure windows are on the same chromosome
     if (dat2$chrom[j] == dat2$chrom[j-1]){
       
       # if the lower CI of this window is LESS than the
       # currentUpperCI that is stored from previous window(s)
       # then the windows DO overlap
       if(dat2$lowerCI[j] <= currentUpperCI){ 
         # no update to currentLowerCI
         # if dat2$upperCI[j] > currentUpperCI,
         # then update the currentUpperCI 
         currentUpperCI <- dat2$upperCI[j]
         # and update the CIs stored in the window df
         windowDF$windowUpperCI[currentWindowID] <- currentUpperCI
         #assign same Id to this window
         dat2$windowID[j] <- currentWindowID
         
       }else{
       # if the lower CI of this window is GREATER than the
       # currentUpperCI that is stored from previous window(s)
       # then the windows DO NOT overlap
         currentLowerCI <- dat2$lowerCI[j]
         currentUpperCI <- dat2$upperCI[j]
         # update Id and assign to this window
         currentWindowID <- currentWindowID + 1
         dat2$windowID[j] <- currentWindowID
         
         # update window DF
         windowDF[currentWindowID,] <- rep(NA,3)
         windowDF$windowID[currentWindowID] <- currentWindowID
         windowDF$windowUpperCI[currentWindowID] <- currentUpperCI
         windowDF$windowLowerCI[currentWindowID] <- currentLowerCI
       } # end if statements to compare CI on same chrom
       
     }else{ # if the next window is on a different chromosome
         currentLowerCI <- dat2$lowerCI[j]
         currentUpperCI <- dat2$upperCI[j]
         # update Id and assign to this window
         currentWindowID <- currentWindowID + 1
         dat2$windowID[j] <- currentWindowID
         # update window DF
         windowDF[currentWindowID,] <- rep(NA,3)
         windowDF$windowID[currentWindowID] <- currentWindowID
         windowDF$windowUpperCI[currentWindowID] <- currentUpperCI
         windowDF$windowLowerCI[currentWindowID] <- currentLowerCI
     }
     
   }# end loop through windows
  dat2$windowID=as.character(dat2$windowID + windowIDstartNum - 1)
  windowDF$windowID=as.character(windowDF$windowID + windowIDstartNum - 1)
  dat3 <- left_join(dat2, windowDF, by = "windowID")
  return(list(windowDF=windowDF, windowCalls=dat3))
} # end function

```


Example application of the `collapseWindow()` function
```{r}
head(invdat)

# Run collapseWindow on the start Break point columns
st <- collapseWindow(invdat$inv_startBreakPt_lowerCI,
               invdat$inv_startBreakPt_upperCI, 
               invdat$chrom,
               invdat$mergeKey)
st

# Run collapseWindow on the end Break point columns
# make sure to give a start ID that is greater than the previous IDs
end <- collapseWindow(invdat$inv_endBreakPt_lowerCI,
               invdat$inv_endBreakPt_upperCI, 
               invdat$chrom,
               invdat$mergeKey,
               windowIDstartNum = 10)
end

# create a list of all the Windows
startWindows <- st$windowDF
endWindows <- end$windowDF

# create a list of all individuals with window ID
startWindows_ind <- left_join(invdat, st$windowCalls, by=
                                join_by("mergeKey", "chrom",
                                        "inv_startBreakPt_lowerCI"=="lowerCI",
                                        "inv_startBreakPt_upperCI"=="upperCI"))
# create a list of all individuals with window ID
endWindows_ind <- left_join(invdat, end$windowCalls, by=
                              join_by("mergeKey", "chrom",
                                        "inv_endBreakPt_lowerCI"=="lowerCI",
                                        "inv_endBreakPt_upperCI"=="upperCI"))

# sanity check
if (nrow(startWindows_ind) != nrow(endWindows_ind)){
  print("Error: different number of rows in `startWindows_ind` and `endWindows_ind`")
  break
}
```


```{r}
## The next step is to collect information from invdat_window on how many
## individuals support each window.
## Then store that information in allWindows

# number of individuals supporting start window 

n_ind_start <- rowSums(table(startWindows_ind$windowID, startWindows_ind$indID)>0)
startWindows2 <- left_join(startWindows, 
                         data.frame(n_ind_start, windowID=names(n_ind_start)),
                        by="windowID")


# number of individuals supporting end window
n_ind_end <- rowSums(table(endWindows_ind$windowID, endWindows_ind$indID)>0)
endWindows2 <- left_join(endWindows, 
                         data.frame(n_ind_end, windowID=names(n_ind_end)),
                        by="windowID")
```


```{r}
## The next step is to collect information from invdat_window on
## how many methods support each window
## Then store that information in allWindows

# number of methods supporting start window 
n_methods_start <- rowSums(table(startWindows_ind$windowID, startWindows_ind$method)>0)
n_methods_start
startWindows3 <- left_join(startWindows2, 
                         data.frame(n_methods_start, 
                                    windowID=names(n_methods_start)),
                        by="windowID")

# number of methods supporting end window
n_methods_end <- rowSums(table(endWindows_ind$windowID, endWindows_ind$method)>0)
endWindows3 <- left_join(endWindows2, 
                         data.frame(n_methods_end, 
                                    windowID=names(n_methods_end)),
                        by="windowID")

rm(startWindows, startWindows2, endWindows, endWindows2)
```


 
```{r}
## The next step is to combine the datasets with the number of individuals and methods supported for each breakpoint

str(invdat)
str(startWindows_ind)
str(endWindows_ind)
str(startWindows3)
str(endWindows3)

invdat_full_start <- left_join(startWindows_ind, startWindows3)
str(invdat_full_start)

invdat_full_end <- left_join(endWindows_ind, endWindows3)
str(invdat_full_end)

# sanity check    
if (nrow(invdat_full_start) != nrow(invdat_full_end)){
  print("Error: different number of rows in `invdat_full_start` and `invdat_full_end`")
  break
}
if (sum(which(colnames(invdat_full_end) %in% c("windowID","windowLowerCI","windowUpperCI"))
!= which(colnames(invdat_full_start) %in% c("windowID","windowLowerCI","windowUpperCI")))){
  print("Error in combining datasets, column names don't match")
  break
}

## Rename the columns
renamecols <- which(colnames(invdat_full_start) %in% c("windowID","windowLowerCI","windowUpperCI"))

colnames(invdat_full_start)[renamecols] <- c("windowID_start","windowLowerCI_start","windowUpperCI_start")

colnames(invdat_full_end)[renamecols] <- c("windowID_end","windowLowerCI_end","windowUpperCI_end")

str(invdat_full_start)
str(invdat_full_end)

invdat_full <- left_join(invdat_full_start, invdat_full_end)

# sanity check    
if (nrow(invdat) != nrow(invdat_full)){
  print("Error: different number of rows in `invdat` and `invdat_full`")
  break
}
``` 

```{r}
## The next step is to ID the inversions and produced a filtered set of inversions
## based on minimum criteria

n_ind_min <- 2 # the minimum number of individuals supporting the call
                # this number will vary depending on the investigator's sample size
n_methods_min <- 2 # the minimum number of methods supporting the call
                # this number will vary depending on the number of methods used

invdat_full$inversionID <- paste(invdat_full$windowID_start, invdat_full$windowID_end, sep="-")

# An inversion passes if both breakpoints are supported by the minimum number
# of individuals and methods
invdat_full$PassInv <- (invdat_full$n_ind_start >= n_ind_min & 
                          invdat_full$n_ind_end >= n_ind_min &
                          invdat_full$n_methods_start >= n_methods_min &
                          invdat_full$n_methods_end >= n_methods_min)
# Again, the PASS criteria may take into account other columns, such as a quality
# score from the program
invdat_full 
```

```{r}
# filter the indidivual-level inversion dataframe
# this is the dataframe to analyze for frequencies of inversions across individuals
invdat_filtered <- invdat_full[invdat_full$PassInv,]
invdat_filtered
```


```{r}
# filter the last dataframe to individual inversions
# this is the dataset to analyze for inversion size distribution
inv_list <- select(invdat_filtered, inversionID, 
                   windowID_start, windowLowerCI_start, windowUpperCI_start,
                   windowID_end, windowLowerCI_end, windowUpperCI_end) 

inv_list_final = filter(inv_list,!duplicated(inv_list))
inv_list_final

# find the maximum size of each inversion from the lowest CI of the start
# breakpoint to the highest CI of the end breakpoint
inv_list_final$invSizeMax <-  inv_list_final$windowUpperCI_end - 
                              inv_list_final$windowLowerCI_start

# find the minimum size of each inversion from the highest CI of the start
# breakpoint to the lowest CI of the end breakpoint
inv_list_final$invSizeMin <-  inv_list_final$windowLowerCI_end - 
                              inv_list_final$windowUpperCI_start

par(mfrow=c(1,1))
hist(inv_list_final$invSizeMax, col="cornflowerblue",
     breaks=seq(0, 10000, by=100), main="", xlab="Inversion size",
     border=adjustcolor("grey80",0))
hist(inv_list_final$invSizeMin, col=adjustcolor("cornflowerblue",0.4),
     border=adjustcolor("grey",0),
     breaks=seq(0, 10000, by=100), add=TRUE)
# TO DO: add a legend for the dark colors maximum size distribution
# and light colors minimum size distribution
```
